Towards an Ecological Trait-data Standard

### 2017-11-10 manuscript/manuscript.Rmd
start shortening manuscript: remove definitions of terms into website

---
title: "Towards an Ecological Trait-data Standard"
author: Florian D. Schneider, Malte Jochum, Gaetane LeProvost, Andreas Ostrowski,
  Caterina Penone, Nadja K. Simons
csl: amnat.csl
output:
  pdf_document: default
  word_document:
    keep_md: yes
geometry: margin=1.5in
header-includes: \usepackage{lineno}
fontsize: 12pt
bibliography: div05.bib
---

@@

Given this prospect, there will be no lack of data, but a lack of data standardisation hampering efficient future use of published datasets. 
To cope with this trend towards distributed data sources, data brokering services such as the German Federation on Biological Data (gfbio.org) are likely to gain importance. These services acquire public data from other sources and simplify the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures. One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.

@@

# 3. Common structure of trait datasets

@@

With this paper, we propose a versatile, minimal data structure, as well as a scheme of a defined vocabulary that captures the different degrees of resolution and measurement detail for multiple use cases of trait-based research. We were drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).  
We were particularly considering the work of @kattge11, and @garnier17, as well as @parr16 to ensure compatibility of datasets with major webservices.

Additionally, a set of terms is proposed to facilitate the definition of traits in trait ontologies or a thesaurus of traits (see below for further detail).

The online reference glossary of terms and description of the proposed structure is found at https://ecologicaltraitdata.github.io/TraitDataStandard/ and should be referred to in the metadata of any dataset that employs the data standard as "Schneider et al. 2017 Ecological Traitdata Standard v1.0, DOI: XXXX.xxxx, URL: https://ecologicaltraitdata.github.io/TraitDataStandard/v1.0/". In addition to the versioned online reference, the dataset should also cite this publication for an explanation of the rationale.  
Individual terms should be referenced via their global Identifier, which is hosted by the GFBio Terminology Service and also can be accessed programmatically (i.e. via the Application Programming Interface, API). Where our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.

We also provide computational tools in an R package to help users transfer their original data into the proposed scheme.
The vocabulary proposed will not capture the entire complexity of trait-based research in ecology and evolution. We invite researchers to comment on and contribute to the next iterations of this standard and develop own applications and ontologies.

## core structure for trait data

As described above, longtable formats are the most predictable and compatible format for complex datasets.

If data contain multilayered information on the measurement or occurrence (including geolocation and date information), further columns may be added to the core set (see description for Extensions below).

If taxon information and metadata need to be included, datasets may apply the structure proposed for EOL TraitBank [@parr16], which applies and expands terms of the Darwin Core Archives [DwCA;].

<!--- -TO DO ---->

@@

In the subsequent sections we propose a set of terms for composing a trait dataset, starting with the columns required for our minimal definition of 'core' trait-datasets proposed above. To enable a better standardization of nomenclatures and values, we add columns that contain 'harmonized' entries, which are linked to semantic ontologies and globally valid identifiers, consistent with semantic web standards [@berners-lee01; @wieczorek12; @page08]. 
We furthermore provide extensions to add information on the level of the occurrence, the measurement or fact, and the entire dataset (i.e. metadata).

### core elements of a trait-dataset

The core content of a row in the trait-dataset is the reported measurement or fact for a single observation, which is composed of a value (`traitValue`) and -- for numeric values -- a standard unit (`traitUnit`). To associate the measurement or fact to a clear trait definition, a unique identifier (`traitID`) links each row to a descriptive trait name (`traitName`).

The core data are also keeping a record of the scientific taxon for which the measurement or fact was obtained in the form of a descriptive taxon name (`scientificName`).

For these core values, we highly recommend to use unambiguous and self-explanatory vocabularies for trait names and taxa. However, to have the data compatible with own databases or analytical code, it might be necessary to use abbreviations or identifiers instead. It is indispensible to relate the user-defined vocabulary to a consensus standard of taxon names or or look-up tables of traits.

@@

<!-- The initiatives to standardise traits into consensus terminology are usually formed around a specific research question, methodology, or organism group and therefore vary in the traits that they consider important.  -->
To date, semantic ontologies exist for plant traits [e.g. TOP Plant Trait Ontology , @garnier17], soil invertebrates [T-SITA, @pey14], or biomedical traits of vertebrates [Vertebrate Trait Ontology, @park13] and more use cases (see OBO foundry for further examples). However, for most organism groups and research contexts we attest a deficit of public ontologies with publicly accessible URIs which would be used as `taxonID`.

<!--### Defining own trait thesauri and ontologies-->

@@

The existing thesauri and ontologies differ widely in terms of hierarchical depths and detail, as well as curation efforts and measures for peer-reviewed quality control. More compatibility of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources. 
We encourage hosting such semantic web resources developed for own projects as public resources, e.g. on OBO Foundry, the GFBio Terminology service, or on GitHub. We provide an example thesaurus, that has been developed to capture traits assessed within the German Biodiversity Exploratories project on Github ().

@@

Beyond measurement units or taxon rank resolution, further information might be available that are not recorded in the core data table, but are related to the trait type, the taxon, the individual or specimen, or to the reported fact, measurement or sampling event.

To avoid redundancy and duplication of content, those information would usually be kept in seperate data sheets of a spreadsheet document. Examples are the place were a specimen has been sampled, the measurement method or statistical precision of the reported value, or the literature source from where a species value has been cited. A unique identifier would link to this other datasheet, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`).

@@

Having separate data sheets linked by identifiers has the advantage of tidy datasets, avoiding duplication of information (@wickham14).
However, for standalone dataset publications on a hosting service without such file format recommendations (e.g. Figshare), data authors may want to avoid the expansion into multiple data tables and rather keep all columns in the same file, which is legitimate. 
It also may be safer to add such information directly as own columns of the data table to facilitate an analysis of cofactors and correlations further down the road. This way, if datasets of different source are merged, the information is readily available without the risk of breaking the reference to an external datasheet.
The trade-off is data-consistency and readability vs. avoidance of duplication.

@@

# Computational tools for producing compliant data

>>>>

---
title: "Towards an Ecological Trait-data Standard"
author:
- name: Florian D. Schneider
- name: Malte Jochum
- name: Gaetane LeProvost
- name: Andreas Ostrowski
- name: Caterina Penone
- name: Nadja K. Simons
csl: amnat.csl
output:
  pdf_document:
    fig_caption: yes
    fig_crop: no
    template: pretty.template
  html_document:
    fig_caption: yes
    fig_width: 4
    highlight: null
    keep_md: yes
    theme: united
  word_document:
    fig_caption: yes
    fig_width: 6
bibliography: div05.bib
---

@@

Given this prospect, there will be no lack of data, but a lack of data standardisation hampering efficient future use of published datasets. 
To cope with this trend towards distributed data sources, data brokering services (e.g. the German Federation on Biological Data; gfbio.org)  or database management systems for scientific projects (e.g. BExIS, XX ) are likely to gain importance. These services simplify and direct the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures. One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.

After all, to mobilise the broad wealth of data into structured databases, it requires a certain minimum of information from the data providers. Getting accurate and well-defined input requires improving literacy of ecological researchers on data standards and ontologies.

@@

# 3. The structure of trait datasets

@@

With this paper, we propose a versatile, minimal data structure, as well as a defined vocabulary that captures the different degrees of resolution and measurement detail for multiple use cases of trait-based research. We were drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).  
We were particularly considering the work of @kattge11, and @garnier17, as well as @parr16 to ensure compatibility of datasets with major webservices and existing standards for biodiversity data management.

The online reference glossary of terms and description of the proposed structure is found at https://ecologicaltraitdata.github.io/TraitDataStandard/.

The vocabulary proposed will not capture the entire complexity of trait-based research in ecology and evolution. We invite researchers to comment on and contribute to the next iterations of this standard and develop own applications and ontologies.

## structure for trait data

As described above, tidy longtable formats are the most predictable and compatible format for the core content of trait datasets and should be used for standalone uploads on low-threshold repositories.

To enable a better standardization of nomenclatures and values, the entries of the core dataset should be related to semantic web references using Uniform Resource Identifiers (URIs) keep a duplicate record of accepted taxon names and standard trait names. In these duplicate columns, values should be reported as 'harmonized' entries, which are linked to semantic ontologies and globally valid identifiers, consistent with semantic web standards [@berners-lee01; @wieczorek12; @page08]. 
If data contain multilayered information on the measurement or occurrence (including geolocation and date information), further columns may be added to the core set (see description for Extensions below). 
For reasons of reproducibility and openness, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files ('.csv' or '.txt').

The metadata of any dataset that employs this data structure should refer to the online version as "Schneider et al. 2017 Ecological Traitdata Standard v1.0, DOI: XXXX.xxxx, URL: https://ecologicaltraitdata.github.io/TraitDataStandard/v1.0/". In addition to the versioned online reference, the dataset should also cite this paper for an explanation of the rationale.

In very complex datasets repetition of verbose entries and long character strings in a single data table should be avoided [@wickham14]. Therefore we recommend use of  Darwin Core Archives [DwCA; @robertson09], the structure proposed for GBIF [@gbif17; http://tools.gbif.org/dwca-assistant/] or EOL TraitBank [@parr15; http://eol.org/info/cp_archives]. This also integrates a defined framework for metadata in the Ecological Metadata Language [@knb11], which should be used for describing and linking the data tables.

@@

Column names and, where possible, field entries should follow the defined vocabulary listed at https://ecologicaltraitdata.github.io/TraitDataStandard/. 
Individual terms should be referenced via their global identifier, which is hosted by the GFBio Terminology Service and also can be accessed programmatically (i.e. via the Application Programming Interface, API). Whereever our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.

### Core columns

To describe the minimum core content, where each row is the reported measurement or fact for a single observation, the following columns are required: a value (column `traitValue`) and -- for numeric values -- a standard unit (`traitUnit`); a descriptive trait name (`traitName`);  the scientific taxon for which the measurement or fact was obtained (`scientificName`). 
For these core values, we highly recommend to use unambiguous and self-explanatory vocabularies for trait names and taxa. 
However, to have the data compatible with own databases or analytical code, it might be necessary to use abbreviations or identifiers instead.

### Relate to ontologies

It is indispensible to relate the user-defined vocabulary to a consensus standard of taxon names or or look-up tables of traits. This is acheived by adding globally valid identifiers for taxa (`taxonID`) and trait definitions (`traitID`), complemented by the more readable accepted names (`ScientificNameStd` and `traitNameStd`, respectively), according to a public ontology (e.g. GBIF Backbone Terminology; for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; in TOP Thesaurus, the traitID for 'fruit mass' would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass').  
By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation and may refer to different scientific community standards  [as practiced in TRY; @kattge11]. This redundancy of data enables continuity on the author and data owner side and quality checking and comparability on the data user side. Keeping an additional reference to a kingdom and class (using columns `kingdom`, `class` as described by DWC and suggested for EOL Traitbank) will avoid ambiguities with taxonomic homonyms. 
It is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for filtering purposes for traits recorded at the family or genus level. This information can be used for trait inference of higher taxonomic resolution using hierarchical probabilistic matrix factorisation [@schrodt15].

If no published trait definition is available that can be referenced, trait-datasets should be accompanied by a dataset-specific glossary of traits, or thesaurus. In its simplest form, a trait thesaurus would provide a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17]. See our online reference for recommendations how to publish a minimal Thesaurus compliant with semantic web standards [http://ecologicaltraitdata.github.io/TraitDataStandard/#terms-for-traitlists-a-trait-thesaurus].  We also provide an example thesaurus, that has been developed to capture traits assessed within the German Biodiversity Exploratories project on Github (http://ecologicaltraitdata.github.io/TraitDataList/).

<!--

@@

To date, semantic ontologies exist for plant traits [e.g. TOP Plant Trait Ontology , @garnier17], soil invertebrates [T-SITA, @pey14], or biomedical traits of vertebrates [Vertebrate Trait Ontology, @park13] and more use cases (see OBO foundry for further examples). However, for most organism groups and research contexts we attest a deficit of public ontologies with publicly accessible URIs which would be used as `taxonID`.

### Defining own trait thesauri and ontologies

@@

The existing thesauri and ontologies differ widely in terms of hierarchical depths and detail, as well as curation efforts and measures for peer-reviewed quality control. More compatibility of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources. 
We encourage hosting such semantic web resources developed for own projects as public resources, e.g. on OBO Foundry, the GFBio Terminology service, or on GitHub. We provide an example thesaurus, that has been developed to capture traits assessed within the German Biodiversity Exploratories project on Github (). 
-->

@@

Beyond measurement units or higher taxon information, further information might be available that may not be core data, but are related to the individual or specimen, or to the reported fact, measurement or sampling event. 
<!--
To avoid redundancy and duplication of content, those information would usually be kept in seperate data sheets of a spreadsheet document. Examples are the place were a specimen has been sampled, the measurement method or statistical precision of the reported value, or the literature source from where a species value has been cited. A unique identifier would link to this other datasheet, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`).

@@

Having separate data sheets linked by identifiers has the advantage of tidy datasets, avoiding duplication of information (@wickham14).
However, for standalone dataset publications on a hosting service without such file format recommendations (e.g. Figshare), data authors may want to avoid the expansion into multiple data tables and rather keep all columns in the same file, which is legitimate. 
-->

<!--- WORK HERE!!! -->

It also may be safer to add such information directly as own columns of the data table to facilitate an analysis of cofactors and correlations further down the road. This way, if datasets of different source are merged, the information is readily available without the risk of breaking the reference to an external datasheet.
The trade-off is data-consistency and readability vs. avoidance of duplication.

@@

# 5. Computational tools for producing compliant data
